Adaptation Writ Large:\\ An Essay in Memory of John H. Holland

### 2015-10-21 paper.tex
MM edits
Did significant editing, plus added all references, and .bib file.

In this short essay, we sketch four key, recurrent themes of Holland's
work on adaptive systems: (1) the evolutionary dynamics of ``building
blocks''; (2) learning via credit assignment and rule discovery; (3)
the emergence of internal models; and (4) the generic properties of
complex adaptive systems.  We discuss the role these themes have
played in computer science, and highlight especially the ideas that we
think remain relevant to today's research agendas.

@@

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{RochesterHolland} and his reading of Ronald Fisher's classic
work on integrating genetics with Darwinian selection \cite{Genetical
  Theory of Natural Selection}.  As Holland further read extensively
in evolutionary biology, economics, game theory, and control theory,
he came to recognize that adaptation was the most central concept in
all these fields.  That is, these fields all concern populations of
agents that must continually obtain information from uncertain
changing environments and use it in order to improve their performance
with respect to those environments.  Moreover, Holland recognized that
in systems with {\em adaptive agents} of this kind, there is never a
state of {\emph equilibrium} or a final {\emph optimum} configuration.
Due to the environment's ``perpetual novelty'' (as Holland termed it),
adaptation continues forever in an open-ended way.  Holland's focus on
open-ended, non-equilibrium dynamics was in stark contrast with the
mainstream approach (at the time) in all these fields---the belief
that ``solving for'' stable equilibrium dynamics was the scientific
goal.  Holland's contrary view was that a system in stable equilibrium
is essentially {\emph dead}.

@@

\item {\bf Exploitation Versus Exploration:} Successful adaptation
  requires the right tradeoff between {\emph exploitation}, in which
  tried-and-true building blocks propagate in a population, and {\emph
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  A system that focuses too much on exploitation
  risks never finding better individuals or not being able to adapt to
  changing environments.  Conversely, too much focus on exploration
  means that the system won't leverage the successful building blocks
  that have already been discovered.  There is some optimal balance
  between these two modes of processing.

\end{itemize}

Several of Holland's early papers (e.g.,
\cite{OutlineLogicalTheory,OptimalAllocation}) and his influential
1975 book {\emph Adaptation in Natural and Artificial Systems}
\cite{ANAS} developed a general, formal setting in which these ideas
could be expressed mathematically.

In particular, inspired by Bellman \cite{AdaptiveControlProcesses} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme {\emph
  exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme {\emph
  exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?

In \cite{OptimalAllocationOfTrials} and \cite{ANAS}, Holland derived
an equation for such a strategy.  Let $A$ denote the arm currently
observed to have the higher payoff probability, and $B$ denote the
other arm.  Holland showed that the optimal strategy (the one that
yields maximal payoff, or minimal loss) is for the number of trials
allocated to $A$ to grow slightly faster than an exponential function
of the number of trials allocated to $B$.  Holland then showed that this
result extends naturally to the {\emph multi-armed bandit} case.

@@

It is important to point out again that the framework Holland
developed in \cite{ANAS} was more general than the genetic algorithm;
it was created in order to develop an interdisciplinary theory of
adaptation, one that would inform biology, say, as much as computer
science \cite{ChristiansenFeldmanPaper}.  The later, successful
application of genetic algorithms to real-world optimization and
learning tasks was, for Holland, just icing on the cake.

@@

While genetic algorithms are nowadays often associated with models of
biological evolution, in Holland's view the processes of adaptation he
was studying manifested themselves in many different complex systems.
One system of special interest to Holland was the human mind, and in
particular, its capacity to learn via induction.

In the 1970s and 80s Holland formulated and extensively explored a
particular model of decision-making, action, and inductive learning,
called the {\emph classifier system}
(e.g., \cite{Holland&Reitman1978,EscapingBrittlenessPaper}.  In a
classifier system, a population of ``if-then'' rules interacts with an
environment, and over time is able to learn to improve its performance
via both {\emph credit-assignment} and {\emph rule-discovery}
algorithms.

Like the genetic algorithm, classifier systems can be viewed both as a
model of adaptation and as an artificial-intelligence (AI) method.  At
the time Holland was developing classifier systems, the field of AI
was focused on expert systems, which typically did not learn on their
own, a fundamental deficiency in Holland's view: ``[Expert] systems
are brittle in the sense that they respond appropriately only in
narrow domains, requiring substantial human intervention to compensate
for even slight shifts in domain'' \cite{EscapingBrittleness}. In
contrast, Holland proposed that ``{\emph induction} is the basic, and
perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines.  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Quoted-in-Waldrop-Book-P.182}.

The if-then rules, when activated, compete to post their results on a
shared ``message list'', modeling the system's short-term memory
(again inspired by Hebb's work).  The competition was based on an
bidding process, in which a rule's current strength affected its
ability to bid highly.  Successful rules were gradually strengthened
(and unsuccessful rules weakened) via a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules would pay back part
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.  The idea for the
bucket-brigade algorithm was inspired by the mechanisms of Hebbian
learning and by early work on learning and optimization in economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1957,Samuel1959}). \footnote{Of course similar
  credit-assignment methods have been used extensively in
  reinforcement learning, neural networks, and other areas of modern
  machine learning, but Holland was among the first to propose such
  methods.}

@@

Classifier systems are fairly complex architectures, with learning
occurring at multiple time scales.  Nevertheless, such systems have
been usefully applied to areas as diverse as
poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock
market \cite{PalmerEtAl}.  Modern reinforcement learning methods have
largely taken over as more useful approaches to real-world decision
and control problems, but classifier systems can perhaps be thought of
as an essential ``stage-setting'' method that enabled the development
of later approaches.

John Holland's main interest in classifier systems was how the two
learning mechanisms could work together to create subsets of rules
that accurately model the environment, and thus are able to ``fire''
at appropriate times.  An important part of such internal models is
the notion of a {\emph default hierarchy}---a set of rules that
together form a kind of hierarchical decision structure, going from
most general to most specific.  For example, a general rule might
state that, ``If X is a bird, then X can fly'', whereas a more specifc
one might state, ``If X is a bird with small wings and a large body,
then X cannot fly'' \cite{InductionBook}.  When faced with a bird in
the environment, both rules might compete to post their conclusion,
but the more specific one would be favored if both of its conditions are met.

Holland envisioned that, with a complex environment and enough time
for learning, a classifier system might learn an intricate default
hierarchy of rules, and that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it has learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
Offspring rules, which begin life weaker than do their parents, can
win the competition and influence the system's behavior only when
there are not strong rules whose conditions are satisfied---in other
words when the system does not know what to do.  If their actions
help, they survive; if not they are soon replaced.  Thus, the
offspring do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{HollandSciAmArticle}.

[MM: Default hierarchy emerged in Goldberg's application?  See Holland's discussion.]

\section{Internal Models}

The concept of {\emph internal models} is central to Holland's theory
of adaptive systems.  He posits that all adaptive systems create and
use internal models to survive in their environments, through
anticipation.  Models can be tacit and learned through evolutionary
time, as in the case of bacteria swimming up a chemical gradient, or
explicit and learned over a single lifespan as in the case of
cognitive systems that incorporate experience into internal rule sets
through learning.  In \underline{Induction}, Holland and his
co-authors discuss how such internal models can be learned, without
supervision, by combining environmental inputs with stored knowledge.
A key idea is that a model defines an equivalence relation over a set
of environmental states, together with a set of transition rules,
which are learned over time based on environmental inputs.  Models
that form valid homomorphisms with the environment allow the system to
make accurate predictions.  In his conception, the equivalence classes
are initially very general, say 'moving object' and 'stationary
object,' and over time through experience and learning they are
specialized into more useful and precise subclasses, say 'insect' and
'nest', which form a default hierarchy and are associated with more
specific transition rules.  Although the idea of default hierarchies
is prevalent in many knowledge representation systems (LIST A FEW?),
Holland made two key contributions, first by emphsizing homomorphisms
as a formal way to evaluate model validity, and idea that dates back
to Ross Ashby's \underline{An Introduction to Cybernetics} that
Holland's student, Bernard Ziegler, developed into a formal theory of
computer modeling and simulation.  Today, even though computational
modeling is a major application area of computing, essential to most
natural sciences and engineering disciplines, these early homorphic
theories of modeling stand as the most elegant approach we know of for
characterizing when a model is consistent with the environment and how
an intelligent agent, human or artificial, can update the model to
better reflect reality.

Holland's second key contribution was describing a computational
mechanism by which a cognitive system could iteratively build up a
detailed and hierarchical model of its environment, as Melanie already
discussed above.

Given that Holland believed that the ability to learn and manipulate
internal models was essential for any adapting system, it is no
suprise that he viewed modeling as essential for scientific inquiry.
Today, we use computational models in several distinct ways---as a
tool for analyzing data in statistical models, through simulation as a
method for discovering new knowledge, and as a way to explain or
understand how a system works.  This third use of models was dear to
Holland's heart.  In his view, the key to science was understanding
the mechanisms that cause a system to behave in a certain way, an
aspiration that goes well beyond data fitting methods, which typically
focus only on describing the aggregate behavior of a system.  For
example, a statistical model that describes the boom and bust pattern
of the stock market cannot address the underlying mechanisms that lead
to these cycles, through the collective actions of many many
individual buy/sell decisions.  Similarly, the genetic algorithms for
which he is so famous provide a simple computational framework in
which to explore the dynamics of Darwinian evolution and whether the
basic mechanisms of variation, differential reproduction, and heredity
are sufficient to account for the richness of our natural world.  This
emphasis on exploratory models to build intuitions was an important
theme of Holland's work, and he often quoted Eddington's remark on the
occasion of the first experimental test of Einstein's theory of
relativity: "The contemplation in natural science of a wider domain
than the actual leads to a far better understanding of the actual."
Holland was interested in models that explored basic principles and
mechanisms, even if they didn't make specific predcitions, models that
could show generically how certain behaviors could be produced, for
example how exponential growth might interact with finite resources.
Holland pioneered a style of modeling that has come to be known as
'individual based' or 'agent based,' in which every component of a
system is represented explicitly and has state, e.g., every trader in
a stock market system or every cell in an immune system model, and
each agent had its own behaviors, which it could update (learn) over
time.  These models are often defined over spatial structures, such as
networks or simple grids, to capture the constraints of systems living
under spatial constraints.  An agent-based model, then, encodes a
theory about the mechanisms that are relevant for producting the
behavior of interest.  Similar to expert systems, such models are
especially useful for studying systems that don't have analytic
mathematical descriptions and they can facilitate interdisicplinary
collaborations because the underlying rules can be easily
communicated.  It should be noted that Holland's view of modeling is
by no means typical.  For example, in a textbook on computational
modeling, the authors offer the following definition: “Modeling is the
application of methods to analyze complex, real-world problems in
order to make predictions about what might happen with various
actions.” [Shiflet and Shiflet, 2006].  This sort of perspective
completely rules out the kind of modeling that Holland was most
interested in, namely exploratory modeling.  Similarly, many dispute
that models make any kind of scientific contribution: “Models are
metaphors that explain the world we don’t understand in terms of
worlds we do.  They are merely analogies, provide partial insight,
stand on someone else’s feet.  Theories stand on their own feet, and
rely on no analogies.”  [Emanuel Derman, 2012].  [NOW WE NEED A STRONG
  FINISHING SENTENCE TO RESCUE JHH STYLE MODELING.]

@@

begin{itemize}
\item Evolutionary computation
\item Q Learning
\item Backprop
\item    From Turing nomination: "Holland's machine learning system known as the
   Learning Classifier System (LCS), developed in the early 1980s,
   incorporated a reinforcement learning algorithm known as the bucket
   brigade for non-Markovian environments, anticipating by nearly a
   decade non-Markovian learning algorithms."
\end{itemize}

\begin{itemize}
\item Exploitation versus exploration -- relevance to reinforcement
  learning, other parts of machine learning and optimization.
\item Active learning.  (Two-armed bandit problem.) 
\item On-line learning.  
\end{itemize}

@@

>>>>

In this short essay, we sketch five key, recurrent themes of Holland's
work on adaptive systems: (1) the evolutionary dynamics of ``building
blocks''; (2) learning via credit assignment and rule discovery; (3)
the emergence of internal models; (4) the art of scientific modeling;
and (5) the universal properties of complex adaptive systems.  We
discuss the role these themes have played in computer science, and
highlight especially the ideas that we think remain relevant to
today's research agendas.

@@

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic work
on integrating genetics with Darwinian selection \cite{Fisher1930}.
As Holland further read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain changing environments and use it in
order to improve their performance with respect to those environments.
Moreover, Holland recognized that in systems with {\em adaptive
  agents} of this kind, there is never a state of {\emph equilibrium}
or a final {\emph optimum} configuration.  Due to the environment's
``perpetual novelty'' (as Holland termed it), adaptation continues
forever in an open-ended way.  Holland's focus on open-ended,
non-equilibrium dynamics was in stark contrast with the mainstream
approach (at the time) in all these fields---the belief that ``solving
for'' stable equilibrium dynamics was the scientific goal.  Holland's
contrary view was that a system in stable equilibrium is essentially
{\emph dead}.

@@

\item {\bf Exploitation Versus Exploration:} Successful adaptation
  requires the right tradeoff between {\emph exploitation}, in which
  tried-and-true building blocks propagate in a population, and {\emph
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  A system that focuses too much on exploitation
  risks never finding better individuals or not being able to adapt to
  changing environments.  Conversely, too much focus on exploration
  means that the system won't leverage the successful building blocks
  that have already been discovered.  There is some optimal balance
  between these two modes of processing.
\end{itemize}

Several of Holland's early papers (e.g.,
\cite{Holland1962,Holland1973}) and his influential
1975 book {\emph Adaptation in Natural and Artificial Systems}
\cite{Holland1975} developed a general, formal setting in which these ideas
could be expressed mathematically.

In particular, inspired by Bellman \cite{Bellman1961} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme {\emph
  exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme {\emph
  exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?

In \cite{Holland1973} and \cite{Holland1975}, Holland derived
an equation for such a strategy.  Let $A$ denote the arm currently
observed to have the higher payoff probability, and $B$ denote the
other arm.  Holland showed that the optimal strategy (the one that
yields maximal payoff, or minimal loss) is for the number of trials
allocated to $A$ to grow slightly faster than an exponential function
of the number of trials allocated to $B$.  Holland then showed that this
result extends naturally to the {\emph multi-armed bandit} case.

@@

It is important to point out again that the framework Holland
developed in \cite{Holland1975} was more general than the genetic algorithm;
it was created in order to develop an interdisciplinary theory of
adaptation, one that would inform biology, say, as much as computer
science \cite{Christiansen1998}.  The later, successful
application of genetic algorithms to real-world optimization and
learning tasks was, for Holland, just icing on the cake.

@@

While genetic algorithms are nowadays most often associated with models of
biological evolution, in Holland's view the processes of adaptation he
was studying manifested themselves in many different complex systems.
One system of special interest to Holland was the human mind, and in
particular, its capacity to learn via induction.

In the 1970s and 80s Holland formulated and extensively explored a
particular model of decision-making, action, and inductive learning,
called the {\emph classifier system} (e.g.,
\cite{Holland1977,Holland1986}.  In a classifier system, a population
of ``if-then'' rules interacts with an environment, and over time is
able to learn to improve its performance via both {\emph
  credit-assignment} and {\emph rule-discovery} algorithms.

Like genetic algorithms, classifier systems can be viewed both as 
models of adaptation and as artificial-intelligence (AI) methods.  At
the time Holland was developing classifier systems, the field of AI
was focused on expert systems, which typically did not learn on their
own, a fundamental deficiency in Holland's view: ``[Expert] systems
are brittle in the sense that they respond appropriately only in
narrow domains, requiring substantial human intervention to compensate
for even slight shifts in domain'' \cite{Holland1986}. In
contrast, Holland proposed that ``{\emph induction} is the basic, and
perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines.  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Waldrop1993} (p. 182).

The if-then rules, when activated, compete to post their results on a
shared ``message list'', modeling the system's short-term memory
(again inspired by Hebb's work).  The competition was based on an
bidding process, in which a rule's current strength affected its
ability to bid highly.  Successful rules were gradually strengthened
(and unsuccessful rules weakened) via a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules would pay back part
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.  The idea for the
bucket-brigade algorithm was inspired by the mechanisms of Hebbian
learning and by early work on learning and optimization in economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1961,Samuel1959}). \footnote{Of course similar
  credit-assignment methods have been used extensively in
  reinforcement learning, neural networks, and other areas of modern
  machine learning, but Holland was among the first to propose such
  methods.}

@@

Classifier systems are fairly complex architectures, with learning
occurring at multiple time scales.  Nevertheless, such systems have
been usefully applied to areas as diverse as
poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock
market \cite{Palmer1994}.  Modern reinforcement learning methods have
largely taken over as more useful approaches to real-world decision
and control problems, but classifier systems can perhaps be thought of
as an essential ``stage-setting'' method that enabled the development
of later approaches.

John Holland's main interest in classifier systems was how the two
learning mechanisms could work together to create subsets of rules
that accurately model the environment, and thus are able to ``fire''
at appropriate times.  An important part of such internal models is
the notion of a {\emph default hierarchy}---a set of rules that
together form a kind of hierarchical decision structure, going from
most general to most specific.  For example, a general rule might
state that, ``If X is a bird, then X can fly'', whereas a more specific
one might state, ``If X is a bird with small wings and a large body,
then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
the environment, both rules might compete to post their conclusion,
but the more specific one would be favored if both of its conditions are met.

Holland envisioned that, with a complex environment and enough time
for learning, a classifier system might learn an intricate default
hierarchy of rules, and that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it has learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
Offspring rules, which begin life weaker than do their parents, can
win the competition and influence the system's behavior only when
there are not strong rules whose conditions are satisfied---in other
words when the system does not know what to do.  If their actions
help, they survive; if not they are soon replaced.  Thus, the
offspring do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{Holland1992}.  A detailed
discussion of some real-world applications of classifier systems is
given in \cite{Booker1989}.

\section{Internal Models and Default Hierarchies}

The concept of {\emph internal models} is central to Holland's theory
of adaptive systems.  He posits that all adaptive systems create and
use internal models to prosper in their environments, through
anticipation.  Models can be tacit and learned through evolutionary
time, as in the case of bacteria swimming up a chemical gradient, or
explicit and learned over a single lifespan, as in the case of
cognitive systems that incorporate experience into internal rule sets
through learning.

In \cite{Holland1989}, Holland and his co-authors discuss how such
internal models can be learned, without supervision, by combining
environmental inputs with stored knowledge.  A key idea is that a
model defines sets of equivalence relations over environmental states,
together with a set of transition rules, which are learned over time
based on environmental ``rewards'' (or ``punishments'').  Models that
form valid homomorphisms with the environment allow the system to make
accurate predictions.  In Holland's conception, the equivalence
classes are initially very general, say ``moving object'' and
``stationary object,'' and over time, through experience and learning,
they are specialized into more useful and precise subclasses, say
``insect'' and ``nest''.  As described above, the adaptive system
eventually learns a default hierarchy of rules involving both general
and specific classes.

Although the idea of default hierarchies is prevalent in many
knowledge representation systems (LIST A FEW?), Holland made two key
contributions.  The first was his emphasis of homomorphisms as a formal way to
evaluate model validity.  This idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
%Today, even though computational modeling is a major
%application area of computing, essential to most natural sciences and
%engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of for characterizing when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update the model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism---classifier systems---by which a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment.  The key learning elements of this method, the
bucket-brigade algorithm, combined with a genetic algorithm, as
described above, presaged many of the ideas in modern reinforcement
learning.

\section{The Art of Scientific Modeling}

Given that Holland believed that the ability to learn and manipulate
internal models was essential for any adapting system, it is no
surprise that he viewed modeling as essential for scientific inquiry.

%Today, we use computational models in several distinct ways---as
%tools for analyzing data in statistical models; as methods for
%discovering new knowledge; and as a means to understand how complex
%systems works.  [Steph, I am not clear on the distinction between the
%second and third things here, so I combined them.  Change back if you
%want.]

Today, we use computational models both for \emph{prediction}---by
analyzing data via statistical models---and for \emph{understanding}
how systems work---by probing the effects of hypothesized underlying
mechanisms.  This latter use of models was dear to Holland's heart.
In his view, the key to science was understanding the mechanisms that
cause a system to behave in a certain way, an aspiration that goes
well beyond data fitting methods, which typically focus only on
describing the aggregate behavior of a system.

For example, a statistical model that describes the boom and bust
pattern of the stock market cannot address the underlying mechanisms
that lead to these cycles, through the collective actions of myriad
individual buy/sell decisions.  Similarly, the genetic algorithms for
which Holland is so famous provide a simple computational framework in
which to explore the dynamics of Darwinian evolution and whether the
basic mechanisms of variation, differential reproduction, and heredity
are sufficient to account for the richness of our natural world.

The emphasis on exploratory models to build intuitions was an
important theme of Holland's work, and he often quoted Eddington's
remark on the occasion of the first experimental test of Einstein's
theory of relativity: "The contemplation in natural science of a wider
domain than the actual leads to a far better understanding of the
actual" \cite{Eddington1927}.

Holland was interested in models that explored basic principles and
mechanisms, even if they didn't make specific or detailed predictions.
Such models can show generically how certain behaviors could be produced.
Holland pioneered a style of modeling that has come to be known as
'individual based' or 'agent based,' in which every component of a
system is represented explicitly and has state, e.g., every trader in
a stock market system or every cell in an immune system model.  In
such models, each agent has its own behavior rules, which it can
update (or learn) over time.  In order to capture the constraints of
systems living under spatial constraints, these models are often
defined over spatial structures, such as networks or simple grids.

A given agent-based model encodes a theory about the mechanisms that
are relevant for producing the behavior of interest.  Similar to
expert systems, such models are especially useful for studying systems
that don't have analytic mathematical descriptions.  Agent-based
models can facilitate interdisciplinary collaborations because the
underlying rules can be easily communicated.  The agent-based models
championed by Holland were typically idealized versions of complex
systems and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

It should be noted that Holland's view of modeling is by no means
typical.  For example, in a textbook on computational modeling, the
authors offer the following definition: “Modeling is the application
of methods to analyze complex, real-world problems in order to make
predictions about what might happen with various actions”
\cite{Shiflet2014}.   This sort of perspective completely rules out the
kind of exploratory modeling that Holland was most interested in.

Some researchers dispute that models make any kind of scientific
contribution: "Models are metaphors that explain the world we don't
understand in terms of worlds we do.  They are merely analogies,
provide partial insight, stand on someone else's feet.  Theories stand
on their own feet, and rely on no analogies."  [Emanuel Derman, 2012].
[STEPH, I COULD NOT FIND REFERENCE FOR THIS QUOTE OUTSIDE SFI VIDEO.]

[NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE
  MODELING.]

@@

\begin{itemize}
\item Evolutionary computation
\item Q Learning
\item Backprop
\item    From Turing nomination: "Holland's machine learning system known as the
   Learning Classifier System (LCS), developed in the early 1980s,
   incorporated a reinforcement learning algorithm known as the bucket
   brigade for non-Markovian environments, anticipating by nearly a
   decade non-Markovian learning algorithms."
\item Exploitation versus exploration -- relevance to reinforcement
  learning, other parts of machine learning and optimization.
\item Active learning.  (Two-armed bandit problem.) 
\item On-line learning.  
\end{itemize}

@@

\bibliographystyle{plain}

\bibliography{paper}
